---
title: "IPL Data Analysis & Match Outcome Prediction"
author: "Yukesh"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    highlight: tango
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
```

---

# Introduction

## Project Overview

The Indian Premier League (IPL) is the world's most popular Twenty20
cricket league, played annually in India since 2008. This project
performs a comprehensive data analysis of IPL matches from 2008 to
2024 and builds machine learning models to predict match outcomes.

## Objectives

The key objectives of this project are:

- Explore and visualize IPL match data across 17 seasons
- Identify key patterns in team performance, toss decisions,
  and venue statistics
- Engineer meaningful features from raw match data
- Build and compare three machine learning models to predict
  which team wins a match
- Evaluate model performance and identify the most important
  predictive features

## Dataset Description

Two datasets were used from Kaggle's IPL Complete Dataset:

- **matches.csv** — 1,095 rows, one per match, covering season,
  teams, toss, venue, and result
- **deliveries.csv** — 260,920 rows, one per ball bowled, covering
  batsman, bowler, runs, and dismissals

---

# Setup & Libraries

```{r libraries}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(readr)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(corrplot)
library(pROC)
```

---

# Data Loading & Exploration

```{r load-data}
matches    <- read_csv("../data/matches_clean.csv")
deliveries <- read_csv("../data/deliveries_clean.csv")

cat("Matches loaded   :", nrow(matches), "rows,", ncol(matches), "cols\n")
cat("Deliveries loaded:", nrow(deliveries), "rows,", ncol(deliveries), "cols\n")
```

## Dataset Structure

```{r structure}
glimpse(matches)
```

## Missing Values

```{r missing}
cat("Missing values in matches:\n")
colSums(is.na(matches))
```

---

# Data Cleaning

The following cleaning steps were applied to the raw data:

1. **Season format** standardized from mixed formats to 4-digit year
2. **Team names** unified — 4 inconsistencies fixed
   (e.g. Kings XI Punjab → Punjab Kings)
3. **No-result matches** removed (5 matches with no winner)
4. **Missing cities** filled from venue names
5. **Irrelevant columns** dropped (method, umpire1, umpire2)
6. **Over numbering** corrected from 0–19 to 1–20
7. **Toss impact column** added

```{r cleaning-summary}
cat("Final cleaned dataset dimensions:\n")
cat("Matches   :", nrow(matches), "rows,", ncol(matches), "cols\n")
cat("Deliveries:", nrow(deliveries), "rows,", ncol(deliveries), "cols\n")
```

---

# Exploratory Data Analysis

## Most Successful Teams

```{r chart1-team-wins, fig.cap="Total IPL wins by each team across all seasons"}
team_wins <- matches %>%
  group_by(winner) %>%
  summarise(wins = n()) %>%
  arrange(desc(wins))

ggplot(team_wins, aes(x = reorder(winner, wins), y = wins, fill = wins)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "#56B4E9", high = "#E69F00") +
  labs(
    title = "Total IPL Wins by Team (2008–2024)",
    x = "Team", y = "Number of Wins"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

**Key Finding:** Mumbai Indians lead with 144 wins, followed closely
by Chennai Super Kings (138) and Kolkata Knight Riders (131).
These three teams form the "Big Three" of IPL.

## Toss Decision Trend

```{r chart2-toss-trend, fig.cap="How toss decisions have changed across IPL seasons"}
toss_trend <- matches %>%
  group_by(season, toss_decision) %>%
  summarise(count = n(), .groups = "drop")

ggplot(toss_trend,
       aes(x = season, y = count,
           color = toss_decision, group = toss_decision)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(values = c("bat" = "#E69F00", "field" = "#0072B2")) +
  labs(
    title  = "Toss Decision Trend Across IPL Seasons",
    x = "Season", y = "Number of Matches", color = "Toss Decision"
  ) +
  theme_minimal(base_size = 13)
```

**Key Finding:** Teams consistently prefer fielding first — 704
matches (64%) vs batting first (391 matches, 36%). This reflects
T20 cricket's chase-friendly nature.

## Does Toss Winning Help?

```{r chart3-toss-impact, fig.cap="Relationship between winning the toss and winning the match"}
toss_impact <- matches %>%
  group_by(toss_winner_won) %>%
  summarise(count = n()) %>%
  mutate(percentage = round(count / sum(count) * 100, 1))

ggplot(toss_impact,
       aes(x = toss_winner_won, y = count, fill = toss_winner_won)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(percentage, "%")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c("Yes" = "#009E73", "No" = "#D55E00")) +
  labs(
    title = "Does Winning the Toss Lead to Winning the Match?",
    x = "Toss Winner Also Won?", y = "Number of Matches"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

**Key Finding:** Toss winners win only 50.8% of the time — barely
above a coin flip. This confirms that the toss has minimal impact
on match outcomes in modern IPL cricket.

## Batting First vs Chasing

```{r chart4-bat-chase, fig.cap="Distribution of wins by batting first versus chasing"}
win_type <- matches %>%
  filter(result %in% c("runs", "wickets")) %>%
  group_by(result) %>%
  summarise(count = n()) %>%
  mutate(
    label      = ifelse(result == "runs",
                        "Won Batting First", "Won Chasing"),
    percentage = round(count / sum(count) * 100, 1)
  )

ggplot(win_type, aes(x = "", y = count, fill = label)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(label, "\n", percentage, "%")),
            position = position_stack(vjust = 0.5),
            size = 5, fontface = "bold") +
  scale_fill_manual(
    values = c("Won Batting First" = "#E69F00",
               "Won Chasing"       = "#56B4E9")) +
  labs(title = "IPL Matches: Batting First vs Chasing Wins") +
  theme_void(base_size = 13) +
  theme(legend.position = "none")
```

## Top 10 Run Scorers

```{r chart5-batsmen, fig.cap="All-time top 10 run scorers in IPL history"}
top_batsmen <- deliveries %>%
  group_by(batter) %>%
  summarise(total_runs = sum(batsman_runs)) %>%
  arrange(desc(total_runs)) %>%
  slice(1:10)

ggplot(top_batsmen,
       aes(x = reorder(batter, total_runs),
           y = total_runs, fill = total_runs)) +
  geom_col() +
  coord_flip() +
  geom_text(aes(label = total_runs),
            hjust = -0.1, size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#56B4E9", high = "#D55E00") +
  labs(
    title = "Top 10 Run Scorers in IPL History",
    x = "Batsman", y = "Total Runs"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, max(top_batsmen$total_runs) * 1.1)
```

**Key Finding:** Virat Kohli leads with 8,014 runs — 1,245 runs
ahead of second-placed Shikhar Dhawan (6,769).

## Top 10 Wicket Takers

```{r chart6-bowlers, fig.cap="All-time top 10 wicket takers in IPL history"}
top_bowlers <- deliveries %>%
  filter(is_wicket == 1) %>%
  filter(!dismissal_kind %in%
           c("run out", "retired hurt", "obstructing the field")) %>%
  group_by(bowler) %>%
  summarise(wickets = n()) %>%
  arrange(desc(wickets)) %>%
  slice(1:10)

ggplot(top_bowlers,
       aes(x = reorder(bowler, wickets),
           y = wickets, fill = wickets)) +
  geom_col() +
  coord_flip() +
  geom_text(aes(label = wickets),
            hjust = -0.1, size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#56B4E9", high = "#009E73") +
  labs(
    title = "Top 10 Wicket Takers in IPL History",
    x = "Bowler", y = "Total Wickets"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, max(top_bowlers$wickets) * 1.1)
```

**Key Finding:** Yuzvendra Chahal leads with 205 wickets,
confirming that leg-spin bowling dominates T20 cricket.

## Matches Per Season

```{r chart7-seasons, fig.cap="Growth of IPL season size over the years"}
season_matches <- matches %>%
  group_by(season) %>%
  summarise(total_matches = n())

ggplot(season_matches,
       aes(x = factor(season), y = total_matches, fill = total_matches)) +
  geom_col() +
  geom_text(aes(label = total_matches),
            vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#56B4E9", high = "#E69F00") +
  labs(
    title = "Number of IPL Matches Played Per Season",
    x = "Season", y = "Total Matches"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

## Top 10 Venues

```{r chart8-venues, fig.cap="Most frequently used IPL venues"}
top_venues <- matches %>%
  group_by(venue) %>%
  summarise(matches_hosted = n()) %>%
  arrange(desc(matches_hosted)) %>%
  slice(1:10)

ggplot(top_venues,
       aes(x = reorder(venue, matches_hosted),
           y = matches_hosted, fill = matches_hosted)) +
  geom_col() +
  coord_flip() +
  geom_text(aes(label = matches_hosted),
            hjust = -0.1, size = 4, fontface = "bold") +
  scale_fill_gradient(low = "#56B4E9", high = "#CC79A7") +
  labs(
    title = "Top 10 IPL Venues by Matches Hosted",
    x = "Venue", y = "Matches Hosted"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  ylim(0, max(top_venues$matches_hosted) * 1.15)
```

**Key Finding:** Eden Gardens, Kolkata has hosted the most IPL
matches (77), followed by Wankhede Stadium, Mumbai (74).

---

# Feature Engineering

Eleven features were engineered from the raw data for the
prediction model:

```{r feature-engineering}
# Rebuild all features
team_win_rate <- matches %>%
  pivot_longer(cols = c(team1, team2),
               names_to = "role", values_to = "team") %>%
  group_by(team) %>%
  summarise(
    total_matches = n(),
    total_wins    = sum(winner == team),
    win_rate      = round(total_wins / total_matches, 3)
  ) %>%
  arrange(desc(win_rate))

cat("Team Win Rates:\n")
print(team_win_rate)
```

```{r build-model-data}
all_teams     <- sort(unique(c(matches$team1, matches$team2)))
team_encoding <- setNames(1:length(all_teams), all_teams)
venue_counts  <- matches %>%
  group_by(venue) %>%
  summarise(venue_match_count = n())

model_features <- read_csv("../data/model_features.csv") %>%
  mutate(team1_wins = factor(team1_wins,
                             levels = c("Team2_Wins", "Team1_Wins")))

cat("\nModel features shape:",
    nrow(model_features), "rows,",
    ncol(model_features), "cols\n")

cat("\nTarget variable distribution:\n")
print(round(prop.table(table(model_features$team1_wins)) * 100, 1))
```

---

# Machine Learning Models

## Train-Test Split

```{r split}
set.seed(42)
train_index <- createDataPartition(model_features$team1_wins,
                                   p = 0.80, list = FALSE)
train_data  <- model_features[ train_index, ]
test_data   <- model_features[-train_index, ]

cat("Training rows:", nrow(train_data), "\n")
cat("Testing rows :", nrow(test_data),  "\n")
```

## Cross Validation Setup

```{r cv-setup}
train_control <- trainControl(
  method          = "cv",
  number          = 10,
  savePredictions = TRUE,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary
)
```

## Model 1 — Logistic Regression

```{r model-lr}
set.seed(42)
model_lr <- train(
  team1_wins ~ ., data = train_data,
  method = "glm", family = "binomial",
  trControl = train_control, metric = "ROC"
)
pred_lr <- predict(model_lr, newdata = test_data)
cm_lr   <- confusionMatrix(pred_lr, test_data$team1_wins,
                            positive = "Team1_Wins")
acc_lr  <- round(cm_lr$overall["Accuracy"] * 100, 2)
cat("Logistic Regression Accuracy:", acc_lr, "%\n")
```

## Model 2 — Decision Tree

```{r model-dt}
set.seed(42)
model_dt <- train(
  team1_wins ~ ., data = train_data,
  method = "rpart", trControl = train_control,
  metric = "ROC", tuneLength = 10
)
pred_dt <- predict(model_dt, newdata = test_data)
cm_dt   <- confusionMatrix(pred_dt, test_data$team1_wins,
                            positive = "Team1_Wins")
acc_dt  <- round(cm_dt$overall["Accuracy"] * 100, 2)
cat("Decision Tree Accuracy:", acc_dt, "%\n")
```

```{r decision-tree-plot, fig.cap="Decision Tree structure for IPL match prediction"}
rpart.plot(
  model_dt$finalModel,
  type          = 4,
  extra         = 108,
  box.palette   = "RdYlGn",
  shadow.col    = "gray80",
  fallen.leaves = TRUE,
  main          = "IPL Match Outcome — Decision Tree"
)
```

## Model 3 — Random Forest

```{r model-rf}
set.seed(42)
model_rf <- train(
  team1_wins ~ ., data = train_data,
  method = "rf", trControl = train_control,
  metric = "ROC", tuneLength = 5, ntree = 500
)
pred_rf <- predict(model_rf, newdata = test_data)
cm_rf   <- confusionMatrix(pred_rf, test_data$team1_wins,
                            positive = "Team1_Wins")
acc_rf  <- round(cm_rf$overall["Accuracy"] * 100, 2)
cat("Random Forest Accuracy:", acc_rf, "%\n")
```

---

# Model Evaluation

## Accuracy Comparison

```{r model-comparison, fig.cap="Accuracy comparison across all three models"}
model_comparison <- data.frame(
  Model    = c("Logistic Regression", "Decision Tree", "Random Forest"),
  Accuracy = c(acc_lr, acc_dt, acc_rf),
  Sensitivity = c(
    round(cm_lr$byClass["Sensitivity"] * 100, 2),
    round(cm_dt$byClass["Sensitivity"] * 100, 2),
    round(cm_rf$byClass["Sensitivity"] * 100, 2)
  ),
  Specificity = c(
    round(cm_lr$byClass["Specificity"] * 100, 2),
    round(cm_dt$byClass["Specificity"] * 100, 2),
    round(cm_rf$byClass["Specificity"] * 100, 2)
  )
)

print(model_comparison)

ggplot(model_comparison,
       aes(x = reorder(Model, Accuracy),
           y = Accuracy, fill = Model)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = paste0(Accuracy, "%")),
            vjust = -0.5, size = 5, fontface = "bold") +
  scale_fill_manual(values = c(
    "Logistic Regression" = "#0072B2",
    "Decision Tree"       = "#E69F00",
    "Random Forest"       = "#009E73"
  )) +
  labs(
    title = "Model Accuracy Comparison",
    x = "Model", y = "Accuracy (%)"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, 100)
```

## Feature Importance

```{r feature-importance, fig.cap="Most important features according to Random Forest"}
importance_df <- varImp(model_rf)$importance %>%
  as.data.frame() %>%
  rownames_to_column("feature") %>%
  arrange(desc(Overall))

ggplot(importance_df,
       aes(x = reorder(feature, Overall),
           y = Overall, fill = Overall)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "#56B4E9", high = "#E69F00") +
  geom_text(aes(label = round(Overall, 1)),
            hjust = -0.1, size = 4, fontface = "bold") +
  labs(
    title = "Feature Importance — Random Forest",
    x = "Feature", y = "Importance Score"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none") +
  ylim(0, max(importance_df$Overall) * 1.15)
```

## ROC Curve

```{r roc-curve, fig.cap="ROC Curve for Random Forest model"}
library(pROC)
pred_rf_prob <- predict(model_rf, newdata = test_data, type = "prob")
roc_rf <- roc(
  response  = test_data$team1_wins,
  predictor = pred_rf_prob[, "Team1_Wins"],
  levels    = c("Team2_Wins", "Team1_Wins")
)
plot(roc_rf,
     col             = "#009E73",
     lwd             = 2.5,
     main            = paste("ROC Curve — Random Forest | AUC =",
                             round(auc(roc_rf), 3)),
     print.auc       = TRUE,
     auc.polygon     = TRUE,
     auc.polygon.col = "#009E7330")
```

---

# Key Findings & Conclusions

## Summary of Findings

| Finding | Insight |
|---|---|
| Most wins | Mumbai Indians — 144 wins |
| Best win rate | Gujarat Titans — 62.2% |
| Toss impact | Only 50.8% — nearly random |
| Fielding first preference | 64% of toss winners chose to field |
| Top batsman | Virat Kohli — 8,014 runs |
| Top bowler | YS Chahal — 205 wickets |
| Most used venue | Eden Gardens — 77 matches |
| Best model | Decision Tree — 55.50% accuracy |
| Top feature | Win rate difference — Score 100 |

## Model Performance Discussion

The best model achieved **55.50% accuracy** using only pre-match
features. While this may appear modest, it represents a meaningful
improvement over the 50% baseline (random guessing) for the
following reasons:

- Cricket is inherently unpredictable — upsets are frequent
- Only 11 pre-match features were used — no live ball-by-ball data
- Professional cricket analytics systems use real-time data and
  achieve 80–85% accuracy using in-match features like current
  score and wickets fallen

## Limitations

1. **No live match data** — only pre-match features used
2. **Venue name mismatches** — some newer venues not found
   in dataset (e.g. Narendra Modi Stadium)
3. **Retired teams** — Pune Warriors and Kochi Tuskers Kerala
   have very few matches, affecting their win rate reliability
4. **Team composition changes** — same team name across years
   can have very different squads

## Future Improvements

1. Add **ball-by-ball live features** like powerplay score and
   wickets to significantly improve accuracy
2. Use **player availability** and recent form as features
3. Apply **deep learning** models like LSTM for sequence-based
   match prediction
4. Build a **Shiny web app** for interactive match prediction

---

# Conclusion

This project successfully performed end-to-end IPL data analysis
covering data loading, cleaning, exploratory analysis, feature
engineering, and machine learning model building. Three models
were trained and compared — Decision Tree emerged as the best
performer with **55.50% accuracy**.

The analysis revealed that **team historical strength** (win rate
difference) and **venue familiarity** are the strongest predictors
of match outcomes, while toss results have minimal impact. Virat
Kohli and Yuzvendra Chahal stand out as the greatest individual
performers in IPL history.

The live match prediction function successfully predicts outcomes
for any combination of IPL teams, providing probability scores
alongside the predicted winner.

---

*Report generated using R `r getRversion()` |
Subject: R for Data Science*